\texttt{ensmallen}: a generic C++ library for fast optimization

### 2018-10-14 doc/mlsys_paper/paper.tex
Significant updates to paper, including some experimental results.

\documentclass{article}
\usepackage{nips_2018}
\usepackage{amsmath}
\usepackage{amsthm}
\usepackage{array}
\usepackage{booktabs}
\usepackage{graphicx}
\usepackage{wasysym}
\usepackage{tabularx}
\usepackage{threeparttable}
\usepackage[newfloat]{minted}
\usepackage{caption}
\usepackage{parcolumns}

@@

The ubiquity of optimization in machine learning algorithms highlights the need
for robust and flexible implementations of optimization algorithms, which is
what drives us to introduce {\tt ensmallen}, a generic C++ optimization toolkit
that contains a wide variety of optimization techniques for different types of
objective functions.  Through the use of template metaprogramming, {\tt
ensmallen} is able to generate efficient code that can help with the
computational intensity encountered with many machine learning algorithms.

@@

\noindent for any function $f(x)$ that takes some vector input $x$.  Ideally,
the user should simply need to specify what $f(x)$ is (or provide an
implementation of a function that can compute $f(x)$).  But it is impossible to
implement something both so generic and fast---for instance, gradient-based
approaches converge far more quickly than gradient-free approaches (in general),
so we must design an abstraction that is able to simultaneously generalize to
many problem types, as well as take advantage of accelerations and
optimizations.

@@

        & & \rot{has framework}

@@

@@

{\it has framework} the library has some kind of generic

@@

Therefore, the best we can hope to achieve is to maximize the flexibility
available, so that a user can easily implement a function $f(x)$ and have it
work with as many optimizers as possible.  For this, {\tt ensmallen} depends on
C++ policy-based design.  When implementing an objective function to be
optimized, a user can implement only a few methods and we can use C++ template
metaprogramming to check that the given functions match the requirements of the
optimizer that is being used.  When implementing an optimizer, a user can assume
that the given function to be optimized meets the required assumptions of the
optimizers, and encode those requirements.

@@

As an example of usage, let us consider the linear regression objective
function\footnote{Just for simplicity, we ignore the bias term.  It can be
rederived by taking $x^*_i = (x_i 1)$.}.  Given some dataset $X \in
\mathcal{R}^{n \times d}$ and associated responses $y \in \mathcal{R}^n$, the
model of linear regression is to assume that $y_i = \theta x_i$ for each data
point and response $(x_i, y_i)$.  To fit this model $\theta \mathcal{R}^d$ best
to the data, we wish to minimize

\begin{equation}
\operatorname{argmin}_\theta \sum_{i = 1}^n (y_i - \theta x_i)^2 =
\operatorname{argmin}_\theta \| y - \theta X \|^2.
\end{equation}

@@

\begin{equation}
f'(\theta) = \sum_{i = 1}^n 2 x_i (y_i - \theta x_i) = 2 X (y - \theta X).
\end{equation}

@@

  // Compute the objective function.
  double Evaluate(const arma::mat& theta)
  {
    return arma::accu((y - theta * X) % (y - theta * X));
  }

  // Compute the gradient and store in 'gradient'.
  void Gradient(const arma::mat& theta, arma::mat& gradient)
  {
    gradient = 2 * X * (y - theta * X);
  }

@@

\vspace*{-0.4em}
% I guess to save some space we should group them.
\begin{itemize} \itemsep -1pt
  \item {\bf SGD variants:} Stochastic Gradient Descent (SGD), Stochastic
      Coordinate Descent (SCD), Parallel Stochastic Gradient Descent (Hogwild!),
      Stochastic Gradient Descent with Restarts (SGDR), SMORMS3, AdaGrad,
      AdaDelta, RMSProp, Adam, AdaMax

@@

  \item {\bf Objective functions:} Neural Networks, Logistic regression,
      Matrix completion, Neighborhood Components Analysis, Regularized SVD,
      Reinforcement learning, Softmax regression, Sparse autoencoders,
      Sparse SVM
\end{itemize}
\vspace*{-0.4em}

@@

It is often the case that when optimizing a function $f(x)$, the computation of
the value $f(x)$ and its derivative $f'(x)$ involve the computation of identical
intermediate results.  Consider the following objective function:

% TODO: need to double check this, and it might be nicer if it was a matrix.
\begin{equation}
f(x) = \| Ax - b \|_F^2.
\end{equation}

For this objective function, the derivative $f'(x)$ is the similar $2A(Ax - b)$,
and both depend on the computation of the matrix term $(Ax - b)$.  If $A$ is a
large matrix, then this may be computationally expensive to compute.  Existing
optimization frameworks do not have an easy way to avoid this duplicate
computation; however, in many cases, an optimization algorithm may need the
values of both $f(x)$ and $f'(x)$ for a given $x$.

Using template metaprogramming in {\tt ensmallen}, we provide an easy (and
optional) way for users to avoid this extra computational overhead.  Instead of
specifying individual {\tt Evaluate()} and {\tt Gradient()} functions, a user
may simply write an {\tt EvaluateWithGradient()} function that returns both the
objective value and the gradient value for an input $x$.

@@

% TODO if space: expand LinearRegressionFunction example to have just an
% EvaluateWithGradient() example.

@@

Let's do some experiments.  I think that maybe we can focus on the following:

\begin{itemize}
  \item An example where the overhead of virtual function calls is
non-negligible---so a function with a very simple Evaluate() and Gradient() but
a very slow convergence with some optimizers (probably SGD... or simulated
annealing?? that could work since SA uses so many iterations).

  \item An example where inlining or compiler optimization can produce
significant benefits.  It's hard for me to think of what this might be.

  \item A machine learning example like logistic regression where
EvaluateWithGradient() avoids a lot of unnecessary computation.  We should avoid
deep neural networks because we're not talking about GPUs here at all.
\end{itemize}

>>>>

\documentclass{article}
\usepackage{nips_2018}
\usepackage{amsmath}
\usepackage{amsthm}
\usepackage{array}
\usepackage{booktabs}
\usepackage{graphicx}
\usepackage{wasysym}
\usepackage{tabularx}
\usepackage{threeparttable}
\usepackage[newfloat]{minted}
\usepackage{caption}
\usepackage{parcolumns}
\usepackage{booktabs}

@@

The ubiquity of optimization in machine learning algorithms highlights the need
for robust and flexible implementations of optimization algorithms, which is
what drives us to introduce {\tt ensmallen}, a generic C++ optimization toolkit
that contains a wide variety of optimization techniques for different types of
objective functions.  Through the use of template metaprogramming, {\tt
ensmallen} is able to generate efficient code that can help with the
demanding computational needs of many machine learning algorithms.

@@

\noindent for any function $f(x)$ that takes some vector or matrix input $x$.  Ideally,
the user should only need to specify what $f(x)$ is (or provide an
implementation of a function that can compute $f(x)$).  But it is impossible to
implement something both so generic and fast---for instance, gradient-based
approaches converge far more quickly than gradient-free approaches (in general),
so we must design an abstraction that is able to simultaneously generalize to
many problem types, as well as take advantage of accelerations and
optimizations.

@@

        & & \rot{unified framework}

@@

        % TODO: Julia
        % TODO: MATLAB paper from MLOSS 2018?

@@

{\it unified framework} the library/language has some kind of generic and unified

@@

Therefore, the best we can hope to achieve is to maximize the flexibility
available, so that a user can easily implement a function $f(x)$ and have it
work with as many optimizers as possible.  For this, {\tt ensmallen} depends on
C++ policy-based design.  When implementing an objective function to be
optimized, a user can implement only a few methods and we can use C++ template
metaprogramming to check that the given functions match the requirements of the
optimizer that is being used.  When implementing an optimizer, a user can assume
that the given function to be optimized meets the required assumptions of the
optimizers, and encode those requirements as \texttt{static\_assert} checks.

For the most common case of a differentiable objective function, the user must
implement only two methods:

\begin{itemize}
  \item \texttt{double Evaluate($x$)}: given coordinates $x$, this function
should return the value of $f(x)$.
  \item \texttt{void Gradient($x$, $g$)}: given coordinates $x$ and a reference
to $g$, set $g = f'(x)$.
\end{itemize}

Or, alternately, they can simply implement a \texttt{EvaluateWithGradient()}
function that computes both $f(x)$ and $f'(x)$ simultaneously, which is useful
in cases where both the objective and gradient depend on similar computations.

The required API for separable differentiable objective functions (i.e. those
that would use an optimizer like SGD) is very similar, except that
\texttt{Evaluate()} and \texttt{Gradient()} (or \texttt{EvaluateWithGradient()})
should operate only on small mini-batches.  The same pattern applies for other
types of objective functions: only a few methods specific to class of objective
function itself must be implemented and then any optimizer may be used.

@@

As an example of usage, let us consider the linear regression objective
function\footnote{Just for simplicity, we ignore the bias term.  It can be
rederived by taking $x^*_i = (x_i, 1)$.}.  Given some dataset $X \in
\mathcal{R}^{n \times d}$ and associated responses $y \in \mathcal{R}^n$, the
model of linear regression is to assume that $y_i = x_i \theta$ for each data
point and response $(x_i, y_i)$.  To fit this model $\theta \in \mathcal{R}^d$ best
to the data, we wish to minimize

\begin{equation}
\operatorname{argmin}_\theta \sum_{i = 1}^n (y_i - x_i \theta)^2 =
\operatorname{argmin}_\theta \| y - X \theta \|^2.
\end{equation}

@@

\begin{equation}
f'(\theta) = \sum_{i = 1}^n -2 x_i (y_i - x_i \theta) = -2 X^T (y - X \theta).
\end{equation}

@@

  // Compute the objective function.
  double Evaluate(const arma::mat& theta)
  {
    return arma::accu((y - X * theta) % (y - X * theta));
  }

  // Compute the gradient and store in 'gradient'.
  void Gradient(const arma::mat& theta, arma::mat& gradient)
  {
    gradient = -2 * X.t() * (y - X * theta);
  }

@@

% I guess to save some space we should group them.
\begin{itemize}
  \item {\bf SGD variants:} Stochastic Gradient Descent (SGD), Stochastic
      Coordinate Descent (SCD), Parallel Stochastic Gradient Descent (Hogwild!),
      Stochastic Gradient Descent with Restarts (SGDR), SMORMS3, AdaGrad,
      AdaDelta, RMSProp, Adam, AdaMax

@@

  \item {\bf Objective functions:} Neural Networks, Logistic regression,
      Matrix completion, Neighborhood Components Analysis, Regularized SVD,
      Reinforcement learning, Softmax regression, Sparse autoencoders,
      Sparse SVM
\end{itemize}

@@

It is often the case that when optimizing a function $f(x)$, the computation of
the value $f(x)$ and its derivative $f'(x)$ involve the computation of identical
intermediate results.  Consider the linear regression objective function from
earlier:

\begin{equation}
f(\theta) = \| y - X\theta \|_F^2.
\end{equation}

For this objective function, the derivative $f'(x)$ is the similar $-2 X^T (y -
X \theta)$, and both depend on the computation of the vector term $(y - X
\theta)$.  If $X$ is a large matrix, then this may be computationally expensive
to compute.  Existing optimization frameworks do not have an easy way to avoid
this duplicate computation; however, in many cases, an optimization algorithm
may need the values of both $f(\theta)$ and $f'(\theta)$ for a given $\theta$.

Using template metaprogramming in {\tt ensmallen}, we provide an easy (and
optional) way for users to avoid this extra computational overhead.  Instead of
specifying individual {\tt Evaluate()} and {\tt Gradient()} functions, a user
may simply write an {\tt EvaluateWithGradient()} function that returns both the
objective value and the gradient value for an input $\theta$.  For our earlier
\texttt{LinearRegressionFunction}, we could use an implementation that only
computes $(y - X \theta)$ once:

\begin{verbatim}
double EvaluateWithGradient(const arma::mat& theta, arma::mat& gradient)
{
  const arma::rowvec v = (y - X * theta); // Cache result.
  gradient = -2 * X.t() * v;
  return arma::accu(v % v); // Take squared norm of v.
}
\end{verbatim}

This method could be provided {\it instead} of {\tt Evaluate()} and {\tt
Gradient()}.

@@

@@

To demonstrate the benefits of these optimizations that {\tt ensmallen} can take
advantage of, we compare {\tt ensmallen}'s performance to a number of other
optimization frameworks, including some that use automatic differentiation.

First, we consider the simple and popular Rosenbrock function: % TODO: cite

\begin{equation}
f([x_1 x_2]) = 100 (x_2 - x_1^2)^2 + (1 - x_1^2).
\end{equation}

To demonstrate the added overhead of other frameworks, we choose to use
simulated annealing, % TODO: cite
a gradient-free optimizer that only uses the objective function itself.
Simulated annealing will call the objective function very many times; for each
simulation we limit the optimizer to $100,000$ objective evaluations.  We
compare four frameworks for this task:

\begin{itemize}
  \item {\tt ensmallen}
  \item {\tt scipy.optimize.anneal}, from scipy 0.14.1 % TODO: cite
  \item {\tt Optim.jl}'s simulated annealing implementation with Julia 1.0.1
% TODO: cite
  \item {\tt samin} from GNU Octave's {\tt optim} package % TODO: cite
\footnote{Another option here might be {\tt simulannealbnd()} from MATLAB's
Global Optimization Toolkit.  However, no license was available for these
simulations.}
\end{itemize}

Table~\ref{tab:rosenbrock_results} shows the average runtime
across 10 trials for $100000$ iterations of simulated annealing with each
implementation.  Of these frameworks, only Julia and {\tt ensmallen} are able to
avoid the function pointer dereference and take advantage of inlining and
related optimizations---and this difference is strongly reflected in the
results.

\begin{table}[t]
\begin{center}
\begin{tabular}{cccc}
\toprule
{\tt ensmallen} & {\tt scipy} & {\tt Optim.jl} & {\tt samin} \\
\midrule
% TODO: these are just single-run results from Marcus' laptop!  We need to do
% 10 and average.
{\bf 0.004s} & 1.044s & 0.021s & 2.920s \\
\bottomrule
\end{tabular}
\end{center}
\caption{Runtimes for $100000$ iterations of simulated annealing with different
frameworks on the simple Rosenbrock function.  Julia code runs do not count
compilation time.}
\label{tab:rosenbrock_results}
\end{table}

Next, we consider the same machine learning example we have been working with
through the paper: linear regression.  For this task we will use the first-order
L-BFGS optimizer, % TODO: cite
so for {\tt ensmallen} we will implement two versions: one with {\tt Evaluate()}
and {\tt Gradient()}, and one with only {\tt EvaluateWithGradient()}.  The code
for each of these is the same as shown earlier.  In addition, we have more
options for Julia: we can implement only the objective function and allow either
% TODO: cite these two
Julia's {\tt Calculus.jl} or {\tt ForwardDiff.jl} package to automatically
compute the gradient.  Therefore, we compare against the following:

\begin{itemize}
  \item {\tt ensmallen}: only {\tt EvaluateWithGradient()} implemented
  \item {\tt ensmallen}-2: both {\tt Evaluate()} and {\tt Gradient()} separately
implemented
  \item {\tt Optim.jl+Calculus.jl}: uses the default {\tt Calculus.jl} for
computing the gradient
  \item {\tt Optim.jl+ForwardDiff.jl}: uses {\tt ForwardDiff.jl} for computing
the gradient
  \item {\tt Optim.jl}: implements the gradient manually
  \item {\tt scipy.optimize.fmin\_l\_bfgs\_b}: implements both objective and
gradient
  \item {\tt bfgsmin} from GNU Octave's {\tt optim} package
\end{itemize}

Results for different data sizes are shown in Table~\ref{tab:lbfgs}.  For each
implementation, L-BFGS was allowed to run for only $10$ iterations and never
converged in fewer iterations.  The datasets trained on were highly noisy random
data with a slight linear pattern---but note that the exact data is not relevant
for our experiments here, only its size.  Runtimes are again reported as the
average across 10 runs.

\begin{table}
\begin{center}
\begin{tabular}{lccccc}
\toprule
{\em algorithm} & $d = 100$, $n = 1k$ & $d = 100$, $n = 10k$ & $d = 100$, $n =
100k$ & $d = 1k$, $n = 100k$ \\
\midrule
% TODO: this was only one trial on Ryan's desktop!
{\tt ensmallen} & {\bf 0.001s} & {\bf 0.006s} & {\bf 0.139s} & {\bf 1.698s} \\
{\tt ensmallen}-2 & 0.002s & 0.008s & 0.204s & 2.497s \\
{\tt Optim.jl+Calculus.jl} & 0.172s & 0.960s & 27.428s & 2535.507s \\
{\tt Optim.jl+ForwardDiff.jl} & 0.553s & 1.162s & 6.665s & 778.250s \\
{\tt Optim.jl} & 0.007s & 0.025s & 0.390s & 4.107s \\
{\tt scipy.optimize.fmin\_l\_bfgs\_b} & 0.002s & 0.010s & 0.198s & 2.022s \\
{\tt bfgsmin} & 0.215s & 1.705s & 15.998s & ??? \\
\bottomrule
\end{tabular}
\end{center}
\caption{Runtimes for linear regression function with several different dataset
sizes.  All Julia runs do not count compilation time.}
\label{tab:lbfgs}
\end{table}

We can see that the use of {\tt EvaluateWithGradient()} yields considerable
speedup of almost 1.5x over the {\tt ensmallen} implementation with both the
objective and gradient implemented separately.  In addition, although the
automatic differentiation support makes it easier for the user to write their
code, it is very clear that the code generated by the automatic differentiator
is far from optimal, with runs that take up to three orders of magnitude longer!
